---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE)       ## set global parameter for echo
options(scipen = 1, digits = 2) ## set rounding defaults
```

## Loading and preprocessing the data
```{r}
  setwd("~/datasciencecoursera/RepData_PeerAssessment1/")

  file    = unz("activity.zip","activity.csv")
  classes = c("numeric","Date","numeric")
  full    = read.csv(file, colClasses = classes)
  df      = full[complete.cases(full),]
```

## What is mean total number of steps taken per day?

A visual presentation of the daily steps can be given by
```{r}
  dailysteps  = tapply(df$steps,df$date,sum)
  hist(dailysteps)
  (meansteps   = mean(dailysteps))
  (medsteps    = median(dailysteps))
```

The median number of steps is `r round(meansteps,0)` and the median is `r round(medsteps,0)`.

## What is the average daily activity pattern?

```{r}
  avgpat  = tapply(df$steps,as.factor(df$interval),mean)
  plot(avgpat,type="l")

```


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
